---
title: "Creel Survey - Stratified Random Sampling"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
This vignette is designed to provide a brief overview of one method to design a creel survey using stratified random sampling. Examples are provided to stratify by time of day, day of week, and multiple access points.  

# Required Packages for this Vignette
Functions used in this chapter require loading the packages shown below.  
```{r, warning=FALSE, message=FALSE}
library(dplyr)      ## for data manipulation
library(tidyr)      ## for data manipulation
library(lubridate)  ## for dates
library(magrittr)   ## for %<>%
library(suncalc)    ## for retrieving sunrise and sunset times
library(ggplot2)    ## for visualizations
```  


# Build Sampling Calendar  
Before we can conduct random sampling, we need to build the sampling calendar which we will randomly draw samples from. This design builds a calendar that classifies holidays as weekends to accurately stratify by day of week. Day length (i.e., sunrise, sunset, and the time between) is also calculated for each day in the sampling calendar.   

### 1. Create a vector of possible sampling dates  
```{r}
## Define start and end dates
start_date <- "2023/03/01"
end_date <- "2023/10/31"
## Create vector based on start and end dates
date_vec <- seq.Date(as.Date(start_date), as.Date(end_date), "days")

## Load function to create a calendar data frame that can be merged with creel data files
source("scripts/calendar.R")

## Build calendar from start and end dates to assign DOW and holidays
cal <- create_calendar(start_date, end_date) %>% 
  ## Add week number
  mutate(week = isoweek(date))

## Display first six rows for example
head(cal)
```  

### 2. Find the sunrise and sunset times for a give location for each date in the vector defined above. 
```{r}
## Define daily sunrise and sunset times
dat <- getSunlightTimes(date_vec, lat = 41.186126, lon = -111.381330, 
                        keep = c("sunrise", "sunset"), tz = "America/Denver") %>% 
  ## Add month variable
  mutate(month = month(date))

## Display first six rows for example
head(dat)
```  


# Examine Day Length During Sampling Period  
A common stratum is time of day (e.g., AM and PM). The rationale for this stratum is explained in the monitoring plan but the figures below provide a visual representation of why this stratum improves feasibility of a creel design.  

```{r}
## Summarize the calendar 
dat_sum <- dat %>% group_by(month) %>% 
  ## find latest sunrise and earliest sunset in each month
  summarize(max_sunrise = max(sunrise),
            min_sunset = min(sunset)) %>% 
  ## build sunrise and sunset variables with common date to calculate length of day
  mutate(sunrise_hour = hour(max_sunrise),
         sunrise_minute = minute(max_sunrise),
         sunrise_time = as.POSIXct(paste0(Sys.Date(), " ", sunrise_hour, ":", 
                                          sunrise_minute), format = "%Y-%m-%d %H:%M"),
         sunset_hour = hour(min_sunset),
         sunset_minute = minute(min_sunset),
         sunset_time = as.POSIXct(paste0(Sys.Date(), " ", sunset_hour, ":", 
                                         sunset_minute), format = "%Y-%m-%d %H:%M")) %>% 
  ## calculate length of day in each month
  mutate(lod = round(sunset_time-sunrise_time, 1))
```

## No Time of Day Stratum  
```{r}
## plot
ggplot(dat_sum, aes(x = month, y = lod)) +
  geom_bar(stat = "identity", fill = "gray70", color = "black") +
  scale_x_continuous(limits = c(2.5, 10.5), breaks = seq(1, 12, 1), expand = c(0, 0.1)) +
  scale_y_continuous(limits = c(0, 16), breaks = seq(0, 16, 2), expand = c(0, 0.02)) +
  labs(y = "Daylength (hours)", x = "Month") + 
  theme_bw() + 
  theme(axis.title.x = element_text(size = 18, margin = margin(10, 0, 0, 0)),
        axis.title.y = element_text(size = 18, margin = margin(0, 10, 0, 0)),
        axis.text.x = element_text(size = 15),
        axis.text.y = element_text(size = 15),
        panel.grid = element_blank(),
        plot.margin = margin(10, 10, 5, 5))
```  


## Add Time of Day Stratum  
```{r}
dat_sum_strat <- dat_sum %>% 
  mutate(losp = lod/2) %>% 
  group_by_all() %>% 
  expand(shift = c("AM", "PM")) %>% 
  ungroup()

ggplot(dat_sum_strat, aes(x = month, y = losp, group = shift, fill = shift)) +
  geom_bar(stat = "identity", color = "black", alpha = 0.8) +
  scale_x_continuous(limits = c(2.5, 10.5), breaks = seq(1, 12, 1), expand = c(0, 0.1)) +
  scale_y_continuous(limits = c(0, 16), breaks = seq(0, 16, 2), expand = c(0, 0.02)) +
  scale_fill_manual(values = c("#1f78b4", "#33a02c")) +
  labs(y = "Length of Sampling Period (hours)", x = "Month") + 
  theme_bw() + 
  theme(axis.title.x = element_text(size = 18, margin = margin(10, 0, 0, 0)),
        axis.title.y = element_text(size = 18, margin = margin(0, 10, 0, 0)),
        axis.text.x = element_text(size = 15),
        axis.text.y = element_text(size = 15),
        legend.position = "inside",
        legend.position.inside = c(0.91, 0.9),
        legend.background = element_rect(fill = "transparent"),
        legend.title = element_blank(),
        legend.text = element_text(size = 15),
        panel.grid = element_blank(),
        plot.margin = margin(10, 10, 5, 5))
```  


# Create Randomized Sampling Calendar
```{r}
## Restrict calendar to four weeks to simplify examples
cal <- cal %>% 
  filter(week %in% 14:17)
```  

## Example 1: Stratified by Time of Day (TOD) and Day of Week (DOW)
```{r}
## set seed to reproduce randomized results
set.seed(254929024)

tod_levels <- c("AM", "PM")

## randomly select two weekdays per week and randomly assign a stratification
## (e.g., AM vs PM) with equal probability
cal_wd_one <- cal %>% filter(dow == "wd") %>% 
  # group by a variable to ensure equal distribution within the levels of 
  # selected variable
  group_by(week) %>% 
  # number of days
  slice_sample(n = 2) %>% 
  ungroup() %>% 
  # randomly assign strata
  mutate(shift = sample(rep(tod_levels, each = ceiling(n()/length(tod_levels))), 
                        size = n(), replace = FALSE))

## randomly select six weekend days per month and randomly assign a stratification 
## (e.g., AM vs PM) with equal probability
cal_we_one <- cal %>% filter(dow == "we") %>% 
  # number of days
  slice_sample(n = 6) %>% 
  ungroup() %>% 
  # randomly assign strata
  mutate(shift = sample(rep(tod_levels, each = ceiling(n()/length(tod_levels))), 
                        size = n(), replace = FALSE))

## Combine randomized calendar
cal_rand_one <- bind_rows(cal_wd_one, cal_we_one) %>% 
  arrange(month, day)
```  
```{r, echo=FALSE}
knitr::kable(cal_rand_one, format = "simple")
```

## Example 3: Stratified by Time of Day (TOD), Day of Week (DOW), and Access Sites
```{r}
## set seed to reproduce randomized results
set.seed(32104934)

site_levels <- c("A", "B", "C")

## randomly select two weekdays per week and randomly assign both strata with 
## equal probability
cal_wd_mult <- cal %>% filter(dow == "wd") %>% 
  # group by a variable to ensure equal distribution within the levels of 
  # selected variable
  group_by(week) %>% 
  # number of days
  slice_sample(n = 2) %>% 
  ungroup() %>% 
  # randomly assign strata (e.g., time of day and access point)
  mutate(shift = sample(rep(tod_levels, each = ceiling(n()/length(tod_levels))), 
                        size = n(), replace = FALSE),
         site = sample(rep(site_levels, each = ceiling(n()/length(site_levels))), 
                       size = n(), replace = FALSE))

## randomly select six weekend days per month and randomly assign both strata with 
## equal probability
cal_we_mult <- cal %>% filter(dow == "we") %>% 
  # number of days
  slice_sample(n = 6) %>% 
  ungroup() %>% 
  # randomly assign strata (e.g., time of day and access point)
  mutate(shift = sample(rep(tod_levels, each = ceiling(n()/length(tod_levels))), 
                        size = n(), replace = FALSE),
         site = sample(rep(site_levels, each = ceiling(n()/length(site_levels))), 
                       size = n(), replace = FALSE))

## Combine randomized calendar
cal_rand_mult <- bind_rows(cal_wd_mult, cal_we_mult) %>% 
  arrange(month, day)
```  
```{r, echo=FALSE}
knitr::kable(cal_rand_mult, format = "simple")
```
